---
title: "Investigating COVID-19 Virus Trends"
output: html_notebook
---

A pneumonia of unknown cause detected in Wuhan, China was first internationally reported from China on 31 December 2019. Today we know this virus as COVID-19, or more casually, as Coronavirus. Since then, the world has been engaged in the fight against this pandemic. Several measures have therefore been taken to "flatten the curve". We have consequently experienced social distancing and many people have passed away as well.

In this project, the dataset used if from [Kaggle](https://www.kaggle.com/lin0li/covid19testing). This dataset was collected between the 20th of January and the 1st of June 2020.

Our analysis tries to provide an answer to this question: **Which countries have had the highest number of positive cases against the number of tests?**

## Understanding the Data

```{r}
library(readr)
# Loading the dataset
covid_df <- read_csv("covid19.csv")
```
```{r}
# Displaing the dimension of the data: 
dim(covid_df)
# Storing the column names in a variable
vector_cols <- colnames(covid_df)
# Displaing the variable vector_cols
vector_cols
# Showing the first few rows of the dataset
head(covid_df)
# Showing a global view of the dataset.
library(tibble)
glimpse(covid_df)

```
The dataset of our study contains daily & cumulative number of COVID-19 tests conducted, number of positive, hospitalized, recovered & death cases reported by country. In details here are the columns in the dataset:

1. `Date`: Date
2. `Continent_Name`: Continent names
3. `Two_Letter_Country_Code`: Country codes
4. `Country_Region`: Country names
5. `Province_State`: States/province names; value is All States when state/provincial level data is not available
6. `positive`: Cumulative number of positive cases reported.
7. `active`: Number of actively cases on that day.
8. `hospitalized`: Cumulative number of hospitalized cases reported.
9. `hospitalizedCurr`: Number of actively hospitalized cases on that day.
10. `recovered`: Cumulative number of recovered cases reported.
11. `death`: Cumulative number of deaths reported.
12. `total_tested`: Cumulative number of tests conducted.
13. `daily_tested`: Number of tests conducted on the day; if daily data is unavailable, daily tested is averaged across number of days in between.
14. `daily_positive`: Number of positive cases reported on the day; if daily data is unavailable, daily positive is averaged across number of days in.

The dataset contains `14` columns and `10,903` rows.


## Isolating the Rows We Need
- Selecting only the rows related to `"All States"` and removing the `Province_State`.

```{r}
library(dplyr)
# Filter the "All States" Province states and remove the `Province_State` column
covid_df_all_states <- covid_df %>% 
  filter(Province_State == "All States") %>% 
  select(-Province_State)

```
- We can remove `Province_State` without loosing information because after the filtering step this column only contains the value `"All States"`.


## Isolating the Columns We Need
- Creating a dataset for the daily columns from `covid_df_all_states` dataframe

```{r}
# Selecting the columns with cumulative numbers
covid_df_all_states_daily <- covid_df_all_states %>% 
  select(Date, Country_Region, active, hospitalizedCurr, daily_tested, daily_positive)
head(covid_df_all_states_daily)
```
## Extracting the Top Ten countries in the number of tested cases

### Summarizing the data based on the `Country_Region` column.
```{r}
covid_df_all_states_daily_sum <- covid_df_all_states_daily %>% 
  group_by(Country_Region) %>% 
  summarise(tested = sum(daily_tested), 
            positive = sum(daily_positive),
            active = sum(active),
            hospitalized = sum(hospitalizedCurr)) %>% 
  arrange(desc(tested))

covid_df_all_states_daily_sum
```
### Taking the top 10
```{r}
covid_top_10 <- head(covid_df_all_states_daily_sum, 10)
covid_top_10
```

## Identifying the Highest Positive Against Tested Cases

### Getting vectors
```{r}
countries <- covid_top_10$Country_Region
tested_cases <- covid_top_10$tested
positive_cases <- covid_top_10$positive
active_cases <- covid_top_10$active
hospitalized_cases <- covid_top_10$hospitalized
```
### Naming vectors
```{r}
names(positive_cases) <- countries
names(tested_cases) <- countries
names(active_cases) <- countries
names(hospitalized_cases) <- countries
```
### Identifying
```{r}
positive_cases
sum(positive_cases)
mean(positive_cases)
positive_cases/sum(positive_cases)
```
```{r}
positive_cases/tested_cases
```
## Conclusion
```{r}
positive_tested_top_3 <- c("United Kingdom" = 0.11, "United States" = 0.10, "Turkey" = 0.08)
```

### Keeping relevant information

```{r}
# Creating vectors
united_kingdom <- c(0.11, 1473672, 166909, 0, 0)
united_states <- c(0.10, 17282363, 1877179, 0, 0)
turkey <- c(0.08, 2031192, 163941, 2980960, 0)
# Creating the matrix
covid_mat <- rbind(united_kingdom, united_states, turkey)
# Naming columns
colnames(covid_mat) <- c("Ratio", "tested", "positive", "active", "hospitalized")
#d Displaying the matrix
covid_mat
```
### Putting all together
```{r}
question <- "Which countries have had the highest number of positive cases against the number of tests?"
answer <- c("Positive tested cases" = positive_tested_top_3)
datasets <- list(
  original = covid_df,
  allstates = covid_df_all_states,
  daily = covid_df_all_states_daily,
  top_10 = covid_top_10
)
matrices <- list(covid_mat)
vectors <- list(vector_cols, countries)
data_structure_list <- list("dataframe" = datasets, "matrix" = matrices, "vector" = vectors)
covid_analysis_list <- list(question, answer, data_structure_list)
covid_analysis_list[[2]]
```

Here we got an answer to our question asked through our analysis.